---
title: "Using BigQuery and SQL to Explore Chicago Crime"
author: "Ilse Paniagua"
output: html_notebook
---

## Setup

```{r}
library(tidyverse)
library(DBI)
library(dbplyr)
library(bigrquery)
```

## Database connection

In this notebook I use Google BigQuery, "Google's fully managed, petabyte scale, low cost analytics data warehouse". Instruction on how to connect to Google BigQuery can be found here:

https://db.rstudio.com/databases/big-query/


```{r}
project <- "ID"
```

First, I set up the connection to the database. I am using the Chicago crime database, which is a BigQuery version of the Chicago crime API.

```{r}
con <- dbConnect(
  bigrquery::bigquery(),
  project = "bigquery-public-data",
  dataset = "chicago_crime",
  billing = project
)
con 
```

These are the tables that are available.

```{r}
dbListTables(con)
```

## SQL and dbplyr

Information on the `crime` table can be found here:

https://cloud.google.com/bigquery/public-data/chicago-crime-data

Writing a first query that counts the number of rows of the `crime` table in the year 2016. The following code chunks expect SQL code.

```{sql connection= con}

SELECT COUNT(*) AS n_rows
FROM `crime` 
WHERE year = 2016

```


Next, I will count the number of arrests grouped by `primary_type` in 2016. Note that is a somewhat similar task as above, with some adjustments on which rows should be considered. Sorting the results, i.e. list the number of arrests in a descending order.

```{sql connection= con}

SELECT primary_type, COUNT(*) AS n_rows
FROM `crime`
GROUP BY primary_type
ORDER BY COUNT(*) DESC

```

I can also use the `date` for grouping. I will count the number of arrests grouped by hour of the day in 2016. You can extract the latter information from `date` via `EXTRACT(HOUR FROM date)`. Which time of the day is associated with the most arrests? 

```{sql connection= con}

SELECT EXTRACT(HOUR FROM date) AS Hour, COUNT (*) AS n_row
FROM `crime`
GROUP BY Hour
ORDER BY COUNT(*) DESC

-- Noon has the most arrests (398820)

```


I focus only on `HOMICIDE` and count the number of arrests for this incident type, grouped by year. Listing the results in descending order.

```{sql connection= con}

SELECT year, COUNT (*) AS n_row
FROM `crime`
WHERE primary_type = 'HOMICIDE'
GROUP BY year
ORDER BY COUNT(*) DESC

```

Finding out which districts have the highest numbers of arrests in 2015 and 2016. That is, count the number of arrests in 2015 and 2016, grouped by year and district. Again, I list the results in descending order (within year).

```{sql connection= con}

SELECT district,
  COUNTIF(year = 2015) AS count_2015,
  COUNTIF(year = 2016) AS count_2016
FROM `crime`
GROUP BY district
ORDER BY COUNTIF(year = 2015) DESC


```

Now, I will compute the difference between the number of arrests in 2016 and 2015 by district. I will order the results such that the district with the highest decrease in arrests comes first.

```{sql connection= con}

SELECT district,
  (COUNTIF(year = 2016)) - 
  (COUNTIF(year = 2015)) AS Difference
FROM `crime`
GROUP BY district
ORDER BY Difference DESC

```

I will demonstrate writing queries from within R via the `DBI` package. I will create a query object that counts the number of arrests grouped by `primary_type` of district 11 in year 2016. The results should be displayed in descending order.

```{r}

sql <- "SELECT primary_type, COUNT(*) AS n_row
        FROM `crime`
        WHERE district = 11 AND year = 2016
        GROUP BY primary_type
        ORDER BY COUNT(*) DESC"
```

Executing the query.

```{r}

dbGetQuery(con, sql)

```

I now write the very same query, now using the `dbplyr` package. For this, I need to first map the `crime` table to a tibble object in R.

```{r}
crime <- tbl(con, "crime")
str(crime)
class(crime)
```

Again, I count the number of arrests grouped by `primary_type` of district 11 in year 2016, now using `dplyr` syntax.

```{r}
crime  %>% select(year, district, primary_type) %>% filter(district==11 & year==2016) %>% group_by(primary_type) %>% summarise(Count=n()) %>% arrange(desc(Count))
```

Counting the number of arrests grouped by `primary_type` and `year`, still only for district 11. Arrange the result by `year`.

```{r}
crime %>% select(primary_type, year, district) %>% group_by(primary_type, year) %>% filter(district==11) %>% summarise(Count=n()) %>% arrange(year, desc(Count))

```

Assigning the results of the query above to a local R object.

```{r}
query <- crime %>% select(primary_type, year, district) %>% group_by(primary_type, year) %>% filter(district==11) %>% summarise(Count=n()) %>% arrange(year, desc(Count))
```

I confirm that I pulled the data to the local environment by displaying the first ten rows of the saved data set.

```{r}

head(query)
```

Closing the connection.

```{r}
dbDisconnect(con)
```
